---
title: Ggplot Short Example
author: Joanna R. Wares
date: "`r format(Sys.time(), '%d %B %Y')`" # be so careful with which quotes you are using
output: html_document
---

```{r, include=FALSE}
library(tidyverse)
library(ggrepel)
#library(smodels)

theme_set(theme_minimal())
options(pillar.min_character_chars = 15)
```


## Plotting with ggplot

We are going to work with the CDC's new cases data. I've downloaded the new cases data for you 
with the first reported value (t=1) being March 1, and put it into a csv file called CDCNewCasesDates.csv. 
You can load it below. The full data can be found [here](https://www.cdc.gov/coronavirus/2019-ncov/cases-updates/cases-in-us.html)

Load the data with the `read_csv()` function. It will be a dataframe with attributes "Date" and "NewCases."
You can see the first few rows of it by using the head() function. Use the read_csv() function to create the dataframe "cases"
```{r cases-data, message=FALSE}
cases <- read_csv(file.path("CDCNewCasesDates.csv"))
head(cases)
```

These are the number of new cases by date in the USA reported early on in the pandemic. 

Now, in the code region below, produce a line plot using  `ggplot()` with `geom_line()` with the date on the x-axis
and new cases on the y-axis. 

Dr. Arnold has some nice notes that we can use to help us found [here](https://statsmaths.github.io/stat289-f20/notes/02_graphics.html)

```{r new-cases-plot, warning=FALSE}
  ggplot(cases, aes(x = Date, y = NewCases)) +
    geom_line()
```

Notice how the dates aren't being recognized as dates. To get them to be recognized as dates, use the `as.Date()` function around the x assignment and replot. 

```{r total-cases, warning=FALSE}
  ggplot(cases, aes(x = as.Date(Date), y = NewCases)) +
    geom_line()

```

This is much better another way to use ggplot is to "pipe" in the data. Here is more information about pipes and there is a lot of r info on this page [https://r4ds.had.co.nz/pipes.html](https://r4ds.had.co.nz/pipes.html). Below is code that produces the same picture as above but with a pipe.
 
Now try to make the same plot but make it blue with a size of 2.

```{r total-cases-plot, warning=FALSE}
cases %>%
  ggplot() +
    geom_line(aes(x = as.Date(Date), y = NewCases),color='blue', size=2)

```


