---
title: "Grammar of Graphics and ggplot2"
author: "Timothy Thornton (tathornt@uw.edu)"
date: '`r Sys.Date()`'
output:
  ioslides_presentation: default
  beamer_presentation: default
subtitle: 'Introduction to the ggplot2 package'
---

```{r echo=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE, 
  collapse = TRUE,
  comment = "#",
  fig.height = 4,
  fig.width = 8,
  fig.align = "center",
  cache = FALSE
) 
```

# Data visualization 

## Data visualization is a critical component of \emph{Data Science} and is used in two main settings
* visualization for *explaining*
* visualization for *exploring*

While common principals apply to both of the above, the aims of the two are different. 

In this lecture we focus primarily on the second type of data visualization, and will introduce tools for exploratory data analysis in R.

# ggplot2 R package  for Data Visualisation  

## ggplot2 R package


* ggplot is a function in the [ggplot2](https://cran.r-project.org/web/packages/ggplot2/index.html) package and is based on **The Grammar of Graphics** by Leland Wilkinson, and the [lattice](https://cran.r-project.org/web/packages/lattice/index.html) package
* ggplot is designed to work in a  *layered fashion*, starting with a layer showing the raw data then adding layers of annotation and statistical summaries 
* The idea is to make the nice features of [lattice](https://cran.r-project.org/web/packages/lattice/index.html) available in a simpler way, and also make it easier to add additional components to the plot (as *layers*, which we talk about later)


## ggplot() function
Let's look at an example **diamonds** data that comes with the [ggplot2](https://cran.r-project.org/web/packages/ggplot2/index.html) package.  But first, let's load the the package to our session using the **library()** function.  If you have not yet installed the ggplot2 package, you should do this first (you only have to do this once).   


```{r echo=FALSE, warnings=FALSE}
#install.packages("ggplot2")
library(ggplot2)
```

Some information about the **diamonds** dataset :

* ~54,000 round diamonds from http://www.diamondse.info/
* Variables:
	+ carat, colour, clarity, cut
	+ total depth, table, depth, width, height
	+ price 

* A question of interest: What is the relationship between *carat* and *price*, and how does it depend on other factors?


## Data on diamonds 

Load the **diamonds** data set, get the dimensions, and look at the first few lines

```{r, warnings=FALSE}
data(diamonds, package="ggplot2")
dim(diamonds)
head(diamonds)
```


## Default R plot of price versus carat

Using the *default settings* with the **plot()** function
```{r}
plot(price~carat, data=diamonds)
```

## Default ggplot price versus diamonds

Using the *default settings* in **ggplot()**

```{r}
ggplot(diamonds, aes(x=carat,y=price)) + geom_point()
```

## A first glance

* The *default* option in **ggplot** looks nicer!
* **ggplot**'s syntax looks weird, especially if you're not very familiar with **lattice**
* **ggplot()** may be slower than **plot()** that comes with base R
* Note: It is possible to manipulate the **plot()** options to get a similar plot (and maybe even a better one), but that would require a lot of extra coding 
* The ggplot *syntax* also makes plotting more  **structured** and  easier to update 

# ggplot syntax

## Using the package ggplot2

Elements of a plot

- data
- aesthetics: mapping of variables to graphical elements
- geom: type of plot structure to use
- transformations: log scale, ...

Additional components 

- layers: multiple geoms, multiple data sets, annotation
- facets: show subsets in different plots
- themes: modifying style



## ggplot syntax

ggplot(diamonds, aes(x=carat,y=price)) + geom_point()

 
The basic concept of a ggplot graphic is to combine different elements into *layers*. Each layer of a ggplotgraphic must have a data set and aesthetic mappings

* **data**: for **ggplot()**, this must be a data frame!
* **aes**: a mapping from the data to the plot; basically the x and y-axes

## ggplot syntax

ggplot(diamonds, aes(x=carat,y=price)) + geom_point()

Layers can also have: 

* a  geom, or a geometric object: defines the overall look of the layer -- is it bars, points, or lines?

* a  stat, or a statistical summary: how should the data be summarized (e.g., binning for histograms, or smoothing to draw regression lines, etc).
* a position: how to handle overlapping points 

When not specified, the defaults are used.

 

##  geom_boxplot
We can use  **geom_boxplot** to create boxplots when one variable is continuous and the other is a factor. 

```{r,}
ggplot(diamonds, aes(x=cut,y=price)) + geom_boxplot()

```


## Changing the aesthetics

You can control the aesthetics of each layer, e.g.  *colour*, *size*, *shape*, *alpha* (opacity) etc.

```{r}
ggplot(diamonds, aes(carat, price)) + geom_point(colour = "blue")
```


## A few more examples
Changing the alpha level

```{r}
ggplot(diamonds, aes(x=carat,y=price)) + geom_point(alpha = 0.2)
```


## A few more examples

Changing the point size

```{r}
ggplot(diamonds, aes(x=carat,y=price)) + geom_point(size = 0.2)

```


## A few more examples

Changing the shape and the point size
```{r}
ggplot(diamonds, aes(x=carat,y=price)) + geom_point(shape = 2,size=0.4)
```

 
## Combining layers
The real power of **ggplot** is its ability to combine layers
```{r}
ggplot(diamonds, aes(x=carat,y=price)) + geom_point(size = 0.2) +
  geom_smooth()
```


## Tranfsormations

In this case (and many other situations) a log transformation may allow for the relationships between variables to be clearer. Can use **coord_trans()**

```{r}
ggplot(diamonds, aes(carat, price)) + geom_point(size = 0.5) +
coord_trans(x = "log10", y = "log10")
```

## Adding information for a third variable
We can color by a *factor* variable (not that it's useful here!)

```{r}
ggplot(diamonds, aes(carat, price, colour=color)) + geom_point() + 
    coord_trans(x = "log10", y = "log10")
```

## Adding information for a third variable
Can also color by a  *continuous* variable (not really useful either!)
```{r}
ggplot(diamonds, aes(carat, price, colour=depth)) + geom_point() + 
    coord_trans(x = "log10", y = "log10")
```




## Adding information for a third variable

In some cases, it may be more useful to get separate plots for each category of the third variable, to understand conditional relationships 

```{r}
ggplot(diamonds, aes(carat, price)) + geom_point() +
  facet_wrap(~color, ncol=4)
```


## Adding information for a third variable

Alternatively, you can use **facet_grid**, which also allows more than 1 conditioning variable (tables of plots)
```{r}
ggplot(diamonds, aes(carat, price)) + geom_point() +
 facet_grid(~color, labeller=label_both)
```


## A final note about syntax
There are actually many ways to get the same plot! The following commands will produce the same plot:

* ggplot(diamonds, aes(price, carat)) + geom_point()
* ggplot() + geom_point(aes(price, carat), diamonds)
* ggplot(diamonds) + geom_point(aes(price, carat))
* ggplot(diamonds, aes(price)) + geom_point(aes(y = carat))
* ggplot(diamonds, aes(y=carat)) + geom_point(aes(x = price))

## A final note about syntax

```{r}
ggplot(diamonds) + geom_point(aes(price, carat))
```

# Additional plots available with ggplot


## Cheat sheet
We are covering only a few of the many plot types that can be greated with the **ggplot2** package

For a more comprehensive view of ggplot2, take a look at the ggplot2 [Cheat sheet](http://www.rstudio.com/wp-content/uploads/2015/12/ggplot2-cheatsheet-2.0.pdf)


## Boxplots

We can summarize univariate distributions using **boxplots**

```{r}
ggplot(diamonds, aes(1, depth)) + geom_boxplot()
```


## Histograms
However, a *histogram* would be a better choice here
```{r}
ggplot(diamonds, aes(depth)) + geom_histogram()
```

Notice the difference in the **aes** call; **boxplot** is really designed for multiple categories!

## Histograms
Tthe *default options* in **histogram** may not be sensible, and you often need to adjust the **binwidth** and **xlim**

```{r}
ggplot(diamonds, aes(depth)) + geom_histogram(binwidth=0.2) + xlim(56,67)
```

## Boxplots with multiple categories
A better use of **boxplot** is when we want to compare distributions of a quantitative variable across categories of a factor variable, as previously discussed

```{r}
ggplot(diamonds, aes(cut, depth)) + geom_boxplot()
```



## Histograms with multiple categories
We can also get multiple histograms, though we need to either display them separately (less useful when comparing)
```{r}
ggplot(diamonds, aes(depth)) + geom_histogram(binwidth = 0.2) + 
    facet_wrap(~cut) + xlim(56, 67)
```

## Overlaying Histograms
Or, you can overlay the historgrams 
```{r}
ggplot(diamonds, aes(depth, fill=cut)) + 
    geom_histogram(binwidth=0.2) + xlim(56,67)
```







